Bayesian non-parametric inference for multivariate peaks-over-threshold models

### 2023-07-07 pointbypoint.tex

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{soul}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{wrapfig}

\newcommand{\comment}[1]{\textcolor{teal}{#1}}
\newcommand{\bruno}[1]{\textcolor{red}{#1}}
\newcommand{\response}[1]{\textcolor{blue}{#1}}

\begin{document}

Dear Professor Jasra,

\vskip 1.5cm

Please consider a revised version of the manuscript `Bayesian non-parametric inference for 
multivariate peaks-over-threshold models''. We have carefully considered the comments made
by the referees, and produced a new version of the paper that we believe shows substantial
improvement with respect to the original version. We notice that addressing the reviewers' 
comments required significant
    effort in redeveloping the simulation study.  To that end, we converted the generating distribution
    for the simulations from mixtures of projected gammas, to mixtures of projected multivariate 
    log-normals so that we are not merely recovering the generating distribution.  Further, the analysis 
    from the simulation study has been made significantly more robust using multiple 
    multiple replicates of the generating distribution.  The results are now reflected in a new figure
    that should make them more easily interpretable.  After a thorough search for available software,
    suitable to perform a comparison with our proposed method, we added two models for 
    comparison: one based on Dirichlet distributions, and one based on pairwise beta distributions.  In
    addition, we have followed the referees' advice of sharpening the presentation, including
    justification, notation and proofs of the stated results.

We believe all concerns raised by the reviewers have been addressed, and we have elaborated our responses
    to each raised issue below.  Thank you for your time and effort in consideration of our manuscript.

% The proof of negative definiteness of our kernel for the energy score has been fixed.  The rotation that
%     previously served as a convenient method of calculation serves in the proof itself to verify negative
%     definiteness.  The ordering of the section has been revised to reflect this.

@@

    % \comment{Hanson is clearly different.  They're using Bernstein polynomials 
    % effectively defined on $\mathbb{S}_1^{d-1}$.  We can demonstrate that 
    % projected gammas built on $\mathbb{S}_{p}$, $p \geq 10$ and $\mathbb{S}_1$ 
    % create \emph{very} different gamma shape parameters, for data created by 
    % the \emph{same} distribution.  I need to check again the other references, 
    % but based on their follow-on comment, I'm guessing the previous point needs 
    % to be addressed before we can properly justify this point.}

    % \bruno{Based on our conversation, you need to look carefully into each one 
    % of these references. it seems that most of them do not develop a PoT 
    % approach. But also, we use a kernel whose natural support is the positive 
    % cone (without using transformations) to build a mixture WRT to a random
    % measure. This provides the flexibility of mixture for a methods that has 
    % computational advantages and scales to moderately large problems, as 
    % illustrated in the paper.}

\item \emph{Numerical accuracy}:  Keeping in mind the computational scope of 
    Statistics and Computing one would hope for more numerical work than what is 
    presented in §5.1. As far as I understand what is presented are single run 
    experiments for each $d$? Monte Carlo evidence on the performance of the 
    procedure would be appreciated. On top of this, the current numerical exercise
    ignores the uncertainty that stems from step 1 of the procedure 
    (i.e. estimating the margins).

\response{To address this concern, we have revised the simulation study to feature the 
    average rise in energy score over baseline, averaged over 10 replications.  
    The simulated datasets are generated as mixtures of multivariate lognormals, 
    projected onto the unit sphere.  The results are now presented in a new set
    of plots in Figure 2. As for the lack of propagation of the uncertainty
    related to the estimation of the Pareto marginals, we acknowledge that this is
    indeed the case. A comprehensive approach is in theory possible, by using the
    full Bayesian formalism. Yet, the positive value function in the marginals 
    introduce data dependent constraints for the parameter space. These are essentially
    impossible to deal with when assuming a flexible model for the angular
    measure. Even for a simple parametric model, the exploration of the posterior
    distribution becomes very challenging in high dimensions. We have a added
    a statement pointing this issue in the section with the concluding remarks.
    }

@@

    \item About the standardization, Eq. (3): please introduce some background. In the paper it is
        introduced as \emph{’the standardization’} and it is unclear what is the relation between this 
        and the definition of the GPD (beginning of Section 2), unless one already knows about marginal
        standardization methods in multivariate EVT. In any case there are several ways to perform 
        marginal standardization in multivariate EVT.

\response{We have rewritten the material at the beginning of Section 3, explicitly mentioning the marginal
        univariate generalized Pareto, and how it is used to estimate the shape and scale parameters, 
        which are then used to standardize the observations.
        }

    \item Section 3.2.1: Please provide some minimal background (or references) on the Dirichlet 
        Process Mixture, in particular about clusters of observations occurring naturally in this 
        model. This is indeed standard in Bayesian Non-parametrics, so there should be a precise 
        reference where it is explained.

\response{To include non-parametric modeling background background to the list of references, in 
        addition to the foundational papers for the DP already included, we added \cite{muller2015} 
        to provide a modern approachable and comprehensive reference,  as well as the more specific
        reference \cite{ascolani2022},  which provides a discussion of robustness of clusters 
        generated under the DP.
        }

@@

\response{We thank the referee for the suggestion to change the layout of the paper. We have
            re-written Section 3 to put a stronger emphasis on the presentation of the
            projected gamma kernel.
            }

        \end{enumerate}

    \item \emph{Unclear Goals}: 
    \begin{enumerate}
        \item There is no proper justification in the paper for the focus on the infinity norm 
        and the associated sphere $\mathbb{S}_{\infty}^{d-1}$. In \cite{ferreira2014}, the focus
        is on the space of bounded processes, and this infinite dimensional sample space is 
        naturally associated with the infinity norm. However in the multivariate case here, the 
        setting is very different and there is no obvious reason why the max norm should be used. 
        Also, by changing the conditioning event, i.e. replacing $\lbrace W \not\leq b_{n}\rbrace$
        (notations of the paper, beginning of Section 2) with 
        $\lbrace \lVert W\rVert_p > t_n\rbrace$, one will obtain a product representation in the 
        limit of the form $\tilde{W} = RV_p$ where $R$ is Pareto and $V_p$ is located on the 
        $\mathcal{L}_p$-sphere. See e.g. one of Resnick’s books, where it is clearly stated that 
        the product form of the limit does not depend on the choice of the norm. I believe that 
        the projected Gamma family on the $L_p$ sphere with a DPM prior is anyways a nice 
        contribution, which is somewhat hidden by this additional projection on the infinity sphere.

\response{We apologize for the lack of clarity of our presentation of the background that supports 
        our proposed model. We have rewritten Section 2 making clear that there are two possible
        approaches to obtaining a multivariate PoT model. One is to assume a max-stability condition 
        that leads to a conditional multivariate Pareto distribution. The other is to assume a 
        condition of regular variation. We follow the former, and use the fact that there are a
        number of representations of the resulting multivariate Pareto. One of them is based on 
        using the infinity norm to represent the random variables that follow such distribution 
        as the product of an angular and a radial component. As for the idea that it is possible 
        to use the $\tilde{W} = RV_p$, where $R$ is a standard Pareto and $V_p$ is located on the 
        $\mathcal{L}_p$-sphere, as a representation of the multivariate Pareto distribution that 
        we term $H$ in the paper, we have  not been able to find any results to justify it for 
        $p\neq \infty$. It is possible to assume regular variation and obtain an equation like
        Equation (1) in the paper for a $p$-norm. That, though, would entail either dealing with,
        or ignoring, the moment constraints and the normalizing constant. Having said that, we
        acknowledge that our model proposes a family of distributions on the $\mathcal{L}_p$-sphere
        that can be useful in its own right. One possible application is that of inference for
        directional data in many dimensions. That is mentioned in the final section of the manuscript.
        }

@@

        \item Another issue is that it is not clear in the present form of the paper, what is the
        added value of fitting this parametric model with the Bayesian framework and the Gibbs 
        sampler, if the output is mainly bivariate densities and extremal dependence coefficients 
        which could be obtained by standard non parametric methods with very low computational cost?

\response{The focus of our paper is to develop a model that is flexible and computationally tractable
        for multivariate PoT extreme value analysis in many dimensions. We discuss and compute the
        properties of the pair-wise extremal coefficients, as it is an important summary of multivariate
        extreme value analysis. Yet, our inferential results are based on the full joint multivariate 
        distribution, and use the whole vector of observations. It is not a pair-wise analysis. In 
        particular, the conditional survival functions considered in Proposition 2 and illustrated in
        Figure 7 make use of the whole joint distribution.
        }
    \end{enumerate}

\item \emph{Comparison with other methods}: The only comparison study displayed in the paper 
    takes place within the suggested model, only with different prior choice. This is not sufficient 
    for performance comparison in a methodological paper. One may e.g. think of comparing with the
    results obtained with the empirical angular measure or smoothed versions of it in terms of 
    extremal dependence and conditional survival probability.

\response{We thank the referee for prompting us to seek a more thorough comparison of our proposed
        model with existing models. To this end we investigated several R and python packages designed 
        to compute empirical angular measure in highly multivariate settings, with the ability to 
        provide predictive uncertainty. We found that most packages implement methods to estimate
        a bivariate angular density,}  \response{including} \verb|ExtremalDep| \response{\citep{ExtremalDep},}
        \response{and} \verb|BAMBI| \response{\citep{BAMBI}. We found that} \verb|evd| \response{\citep{evd} implements a multivariate 
        generalized Pareto density with a  parametric dependence structure, but it is unclear to us 
        how to learn the dependence structure and sample from it. 
        Notably} \verb|BMAmevt| \response{\citep{BMAmevt} implements the 
        pairwise betas model developed in \cite{sabourin2013}, as well as the Dirichlet distribution,
        an example of a multivariate density on $\mathbb{S}_1^{d-1}$.  Our first attempt to use
        it found a bug with its output for distributions with more than 3 dimensions.  Upon 
        conferring with the package maintainer and author, they identified the source of the
        bug and fixed it.  For this reason, we are able to include it as a point of comparison. The
        comparison consists of the energy scores. For the IVT data we also include the computing
        times. Due to the very large nature of the exercise, computations for the simulated examples 
        were performed in a number of different environments, using different number of processors. Thus
        computation times are difficult to compare. For the IVT examples all computations were performed 
        in the same computer, serially. Thus, the comparison of computing times is fair. 
}

@@

    \item \emph{Experiments}:
        \begin{enumerate}
            \item Computational Evidence missing: The convergence of the Gibbs sampler is not 
            discussed in the paper.

\response{We thank the reviewers for bringing this omission to our attention.  We have now
            included a discussion of MCMC convergence among the models in section 5.  In the current
            version of the paper we clarify that we used two different strategies to run the MCMC.
            For the model that have a prior distribution centered around a log-normal we observed 
            that a single chain Metropolis-Hastings algorithm had very slow mixing. Thus we resorted 
            to implementing a parallel tempering approach. This improved the mixing, at the cost 
            of increasing computation times per iteration. In addition, parallel tempering requires
            tuning of the temperature scale, something that can be time-consuming. The models that
            use a prior centered around a product of gammas have substantially better mixing than the
            log-normal centered models, albeit the mixing is not ideal. Yet each iteration is much
            faster, so we adopt a strategy of running the chain longer and doing more thinning, than 
            in the former case. This is now discussed in Section 5. In order to monitor convergence 
            we used plots of the log-posterior density.
            Below are the log-density by iteration plots for the ERA-5 data, on the two 
            unrestricted projected gamma models; left uses the product of gammas centering distribution, 
            right uses the log-normal centering distribution.  The red line indicates the end of the 
            burn-in period, and we only keep every tenth sample after.
            }

            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-G.pdf}
            \includegraphics[width=0.45\textwidth]{images/ERA-5-PG-LN.pdf} \\

            \response{The choice of 40,000 iterations burn-in, and thinning the resulting chain, 
            provides a consistent strategy that produces satisfactory posterior samples for both 
            models.}

            % \bruno{You should include, also, a graph of the autocorrelation functons of these two values.
            % Now, there is something that is bit contradicting, you say you run the PG-G longer and do more
            % thinning and the PG-LN shorter and do less thinning. But in both cases you use 40K and thin
            % every 10. How is that?}

### 2023-07-07 conclusion.tex
prep for submission

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto 
    described in \cite{ferreira2014} to establish a useful representation of 
    its dependence structure, through the distribution of its angular component,
    which is supported on the positive orthant of the unit hypersphere under 
    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
    obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
    our method transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of 
    products of independent gammas, then transforms the predictions back to 
    ${\mathbb S}^{d-1}_\infty$. As ${\mathbb S}_{p}^{d-1}$ converges to 
    ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the proposed 
    resampling to be efficient for large enough $p$. In fact, our exploration 
    of the simulated and real data indicates that the procedure is robust to 
    the choice of moderately large values of $p$. 
    Our method includes two inferential steps. The first consists of the
    estimation of the marginal Pareto distributions. The second consists 
    of the estimation of the angular density. Parameter uncertainty incurred in
    the former is not propagated to the latter. Conceptually, and integrated approach 
    that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
    this leads to posterior distributions with complex data dependent restrictions 
    that  are very difficult to explore, especially in large dimensional setting. In 
    fact, our attempts to fit a simple parametric model for the marginals and the
    angular measure jointly in several dimensions were not successful.

@@

We explored the behavior of the proposed model using simulated data from
    mixtures of projected Gammas with varying degrees of complexity---both 
    in terms of dimensionality and in number of mixture components.  From this,
    we observed that the additional flexibility offered by varying the $\beta$
    term---the rate parameter of the gamma distribution---can offer additional
    model fidelity.  \makenote{However,} this result is not borne out in our 
    real data example, for which in higher dimensions the simpler model is 
    strongly preferred.  Additionally, we observe
    that the choice of the log-normal centering distribution and additional
    flexibility it offers did not translate to additional model fidelity.  
    \makenote{This result is borne out in the real data, where it performed at best 
    comparably to the gamma model.}  
    % This
    % is borne out in the real data, where on the ERA--Interim dataset it performed
    % only comparably, but on ERA--5 it performed significantly worse.
    % {\bf I am not sure that this paragraph is appropriate anymore. I think that
    % the different mixture models considered are pretty much equivalent, so we should
    % favor the one that is easiest and most robust to fit.}

>>>>

\section{Conclusion\label{sec:conclusion}}
In this paper, we have built upon the definition of the multivariate Pareto 
    described in \cite{ferreira2014} to establish a useful representation of 
    its dependence structure, through the distribution of its angular component,
    which is supported on the positive orthant of the unit hypersphere under 
    the $\mathcal{L}_{\infty}$ norm.  Due to the inherent difficulty of 
    obtaining the likelihood of distributions with support on ${\mathbb S}^{d-1}_\infty$ 
    our method transforms the data to ${\mathbb S}_{p}^{d-1}$, fits then using mixtures of 
    products of independent gammas, then transforms the predictions back to 
    ${\mathbb S}^{d-1}_\infty$. As ${\mathbb S}_{p}^{d-1}$ converges to 
    ${\mathbb S}_{\infty}^{d-1}$ as  $p\to\infty$, we expect the proposed 
    resampling to be efficient for large enough $p$. In fact, our exploration 
    of the simulated and real data indicates that the procedure is robust to 
    the choice of moderately large values of $p$. 
    Our method includes two inferential steps. The first consists of the
    estimation of the marginal Pareto distributions. The second consists 
    of the estimation of the angular density. Parameter uncertainty incurred in
    the former is not propagated to the latter. Conceptually, an integrated approach 
    that accounts for all the estimation uncertainty is conceivable. Unfortunately, 
    this leads to posterior distributions with complex data dependent restrictions 
    that  are very difficult to explore, especially in large dimensional settings. In 
    fact, our attempts to fit a simple parametric model for the marginals and the
    angular measure jointly in several dimensions were not successful.

@@

%We explored the behavior of the proposed model using simulated data from
    %mixtures of projected Gammas with varying degrees of complexity---both 
    %in terms of dimensionality and in number of mixture components.  From this,
    %we observed that the additional flexibility offered by varying the $\beta$
    %term---the rate parameter of the gamma distribution---can offer additional
    %model fidelity.  \makenote{However,} this result is not borne out in our 
    %real data example, for which in higher dimensions the simpler model is 
    %strongly preferred.  Additionally, we observe
    %that the choice of the log-normal centering distribution and additional
    %flexibility it offers did not translate to additional model fidelity.  
    %\makenote{This result is borne out in the real data, where it performed at best 
    %comparably to the gamma model.}  
    % This
    % is borne out in the real data, where on the ERA--Interim dataset it performed
    % only comparably, but on ERA--5 it performed significantly worse.
    % {\bf I am not sure that this paragraph is appropriate anymore. I think that
    % the different mixture models considered are pretty much equivalent, so we should
    % favor the one that is easiest and most robust to fit.}

### 2023-07-07 introduction.tex
prep for submission

The reminder of this paper is outlined as follows. 
    Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, 
      detailing the separation of the radial measure from the angular measure.
    Section~\ref{sec:methodology} details our approach for estimating the angular 
      measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ 
      onto unit hyper-spheres defined for $p$-norms. 
    Section~\ref{sec:evaluation} develops criteria for model selection in the support 
      of the angular measure.  
    Section~\ref{sec:results} explores the efficacy of the proposed approach on a set 
      of simulated data, and, acknowledging the relevance of extreme value theory to 
      climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the 
      extremal dependence structure for a measure of water vapor flow in the atmosphere, 
      used for identifying atmospheric rivers.  
    Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.

@@

The approach considered in \cite{rootzen2018} focuses on the limiting
    conditional distribution $H$. An alternative approach consists of assuming
    that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
    for the limiting distribution of $\bm{W}$, implying
    \[
        \lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = 
        \mu(A),
    \]
    where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
    $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
    and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
    \bm{\theta} \in B\})$, which is referred to as the angular measure. After 
    some manipulations, we obtain that
    \begin{equation}
    \label{exponent-mu}
        \lim\limits_{r\rightarrow\infty} 
        \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
            \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
    \end{equation}
    Thus, a model for the exponent measure induces a model for the limiting distribution
    conditional on the observations being above a threshold defined with respect
    to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
    standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
    Inference for the limiting distribution of the exceedances
    needs to account for the normalizing constant in Equation \ref{exponent-mu}, as well
    the moment constraints. In addition, the moment constraints can impose unrealistic
    symmetries in the data distribution.

>>>>

The remainder of this paper is outlined as follows. 
    Section~\ref{sec:multivariatepot} comprises a brief review of multivariate PoT, 
      detailing the separation of the radial measure from the angular measure.
    Section~\ref{sec:methodology} details our approach for estimating the angular 
      measure, based on projecting an arbitrary distribution supported in ${\mathbb R}_+^d$ 
      onto unit hyper-spheres defined by $p$-norms. 
    Section~\ref{sec:evaluation} develops criteria for model selection in the support 
      of the angular measure.  
    Section~\ref{sec:results} explores the efficacy of the proposed approach on a set 
      of simulated data, and, acknowledging the relevance of extreme value theory to 
      climatological events~\citep{jentsch2007,vousdoukas2018,li2019}, estimates the 
      extremal dependence structure for a measure of water vapor flow in the atmosphere, 
      used for identifying atmospheric rivers.  
    Finally, Section~\ref{sec:conclusion} presents our conclusions and discussion.

@@

The approach considered in \cite{rootzen2018} focuses on the limiting
    conditional distribution $H$. An alternative approach consists of assuming
    that regular variation \cite[see, for example,][]{resnick2008extreme} holds 
    for the limiting distribution of $\bm{W}$, implying
    \[
        \lim\limits_{n\rightarrow\infty} n \prob{n^{-1} \bm{W}\in A} = 
        \mu(A),
    \]
    where $\mu$ is termed the exponent measure. $\mu$ has the homogeneity property
    $\mu(tA) = t^{-1}\mu(A)$. Letting $\rho = \|W\|_p, p>0$ 
    and $\bm{\theta} = \bm{W}/\rho$, define  $\Psi(B) = \mu(\{\bm{w} : \rho>1, 
    \bm{\theta} \in B\})$, which is referred to as the angular measure. After 
    some manipulations, we obtain that
    \begin{equation}
    \label{exponent-mu}
        \lim\limits_{r\rightarrow\infty} 
        \text{Pr}\left[\bm{\theta}\in A \rvert \rho>r\right] = 
            \frac{\Psi(A)}{\Psi({\mathbb S}_p^{d-1})} .
    \end{equation}
    Thus, a model for the exponent measure induces a model for the limiting distribution
    conditional on the observations being above a threshold defined with respect
    to their $p$-norm. The constraint that all marginals of $\mu$ correspond to a
    standard Pareto distribution leads to the so called moment constraints on $\Psi$. 
    Inference for the limiting distribution of the exceedances
    needs to account for the normalizing constant in Equation (\ref{exponent-mu}), as well
    the moment constraints. In addition, the moment constraints can impose unrealistic
    symmetries in the data distribution.
